01_Vegetation_cover

### 2022-09-08 Manuscript.Rmd
adjusted introduction after comments

@@

In the past two decades, plant functional traits have become an important tool in ecology and biogeography for the study of consequences of changing plant composition on ecosystem functioning and the effect of environment factors on plant composition [@Hevia_2017]. Recently, considerable interest in the trait-based approach has arisen in palaeoecology as well [@Brussel_2018; @Blaus_2020; @Carvalho_2019; @Reitalu_2015; @Sande_2019]. Applying the trait-based approach could offer a new way of interpreting pollen data and increase our understanding of the drivers of vegetation change on timescales of centuries to millennial. For instance, @Sande_2019 combined functional trait and pollen data to study the effect of climate and human disturbances on vegetation composition in the Amazon on a 7000-year time scale.

Although palaeoecological studies that have applied the trait-based approach show promising results, it is a relatively new approach of which the intricacies are underexplored. Little is known about how representative the pollen record is of plant trait composition. [-include what others have found-.]{.underline} To improve our understanding of the relationship between the trait composition of the vegetation and trait compositions reconstructed from pollen records, we here use surface pollen samples. Surface pollen assemblages approximate vegetation composition in recent years and are commonly used in palaeoecology to verify pollen-vegetation relationships [@Birks_2016]. Here, we use surface samples of 16 Scottish and 26 Swiss sites to test the relationship between reconstructed trait composition from pollen data and vegetation trait composition.

Pollen quantity does not directly represent past plant abundance, taxa can become over- or underrepresented in the pollen record because of differences in pollen production and dispersal. [-something about pollination syndrome and plant function type]{.underline} - To correct for these differences pollen productivity factors can be use, these correction factors down-weight the contribution of dominant taxa. Furthermore, comparing between functional groups based on growth form or pollination syndrome could give insights in to the representativeness of the pollen assemblage to the vegetation. [-something about why pollination syndrome, distance from sampling site]{.underline} [and plant function type would matter to representation -]{.underline}

Perhaps the most challenging in reliably reconstructing past trait is to attribute functional trait data to the taxa found in pollen records, which can rarely be resolved to the species level [@Birks_2020]. Trait variation within taxa can be substantial [@Siefert_2015]. It can be expected that this especially problematic in the case of large, abundant plant families that cannot easily be resolved to a lower taxonomic level, such as the Poaceae family. Also small families where the trait variation is high could be problematic, such as is the case with the Fabaceae family where the height variation can be ...-fold (for instance Ulex shrub height and clover). Vegetation types with a large proportion of taxa with large trait variation might therefore be less reliably reconstructed. One way of dealing with the uncertainty caused by trait variation in pollen taxa is by using Bayesian models. Using a Bayesian approach for modelling trait composition from pollen data allows for the propagation of uncertainty in the taxon level trait reconstruction to the community level.

With this study, we aim to test the validity of using of pollen records for plant trait composition reconstruction. More specifically, we will test the effect of vegetation composition, taxonomic resolution and trait choice on the representativeness of pollen assemblages for functional diversity reconstructions. We examine the effectivity of pollen production factors to improve the pollen - functional diversity relationship. Lastly, we test the representation of plant functional groups based on growth form and pollination syndrome in the functional composition reconstruction.

@@

@@

### Vegetation surveys

@@

### Trait collection

Trait data was collected in the middle ring. For determining plant height 10 flowering of individuals were selected at every site if present, non-flowering individuals were selected if flowering individuals were not available. Five leaf samples per species for specific leaf area (SLA), leaf area (LA), leaf dry matter content (LDMC), leaf carbon and leaf nitrogen were collected in sealable plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbad scanner. Samples were dried at 70 C for 72 hours before measuring dry weight. For measuring leaf carbon and nitrogen samples using ICP-MS were crushed and X.

@@

For reconstructing plant trait composition from pollen data, assumptions need to be made about the species belonging to the pollen taxa in the data set. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name.

### Preparation of pollen data

@@

@@

    Correlations were calculated for all zones and three traits (currently): SLA, LA and plant
    height. Furthermore, four different \"treatments\" were assigned to test what factors could influence representation of trait composition by pollen data: the use of correction factors, the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

@@

## ![](Figures/veg-Family.png)Appendix

@@

>>>>

## Abstract

@@

Functional traits provide a better insight in to the processes that drive biodiversity and how biodiversity affects ecosystem functioning [@Hevia_2017]. While the functioning of modern ecosystems is regularly contingent on past human and environmental factors[@Levis_2018; @Jogiste_2017; @Blonder_2018], the observational duration of trait-based studies is generally short[@Estes_218]. So to extend the time scales of ecosystem functioning research, considerable interest in the trait-based approach has arisen in palaeoecology [@Brussel_2018; @Blaus_2020; @Carvalho_2019; @Reitalu_2015; @Sande_2019]. Applying the trait-based approach in palaeoecological studies could offer a new way of interpreting pollen data and increase our understanding of the drivers of vegetation change on timescales of centuries to millennia. For instance, @Sande_2019 found that on a 7000-year time scale human disturbance superseeds climate effects on plant trait composition in the Amazon.

Although palaeoecological studies that have applied the trait-based approach show promising results, it is a relatively new approach of which the intricacies are underexplored. Little is known about how representative the pollen record is of past plant trait composition. Two main methodological issues arise when attempting to reconstruct functional trait composition from pollen data. First, pollen quantity does not exactly represent past plant abundance. Second, the low taxonomic resolution of the pollen record makes it not straightforward to assign traits to pollen taxa, as trait data is generally collected at the species level.

Taxonomic differences in pollen dispersal and production can cause an over- or underrepresentation of taxa in the pollen record. Wind-pollinated and woody taxa are generally better represented in the pollen record than herbaceous and insect-pollinated taxa as their pollen are produced in larger quantities and dispersed over greater distances. The pollen source area for wind pollinated taxa can therefore be a couple magnitudes greater than that of wind pollinated taxa [\@Meltsov_2011](Bunting 2022?) (extend argument? bunting 2003, bunting 2022). It has therefore been suggested by @Blaus_2020 to treat woody and non-woody taxa separately in the reconstruction of functional composition from pollen data. Another or additional way to correct for taxon specific differences use pollen productivity estimates, these correction factors down-weight the contribution of dominant taxa [@Githumbi_2022].

Perhaps the most challenging in reliably reconstructing past trait is to attribute functional trait data to the taxa found in pollen records. It is rarely possible to resolve the pollen record to the species level [@Birks_2020]. Trait variation within taxa can be substantial depending on the trait and the diversity of the taxon [@Albert_2010; @Prinzing_2008]. It can be expected that this especially problematic in the case of large, abundant plant families that cannot easily be resolved to a lower taxonomic level, such as the Poaceae family. @Blaus_2020 and @Carvalho_2019 proposed that the high species number included in herbaceous pollen taxon could be one of the reasons why herbaceous vegetation was less well represented by pollen based functional diversity reconstructions. In both studies, however, traits values of pollen taxa were calculated by averaging over the available species level trait information. In this study, we use a Bayesian approach for modelling trait composition from pollen data, thereby allowing for the propagation of uncertainty in the taxon level trait reconstruction to the community level.

With this study, we aim to test the validity of using of pollen records for plant trait composition reconstruction. We use surface pollen samples as they approximate vegetation composition in recent years [@Birks_2016]. We examine the effectiveness of pollen production factors to improve the pollen - functional diversity relationship. Furthermore, we test the representation of different plant functional groups by the pollen-based functional trait reconstruction. Laslty, we investigate the effect of taxonomic resolution on the reconstructions.

@@

### Study locations

Here, we use surface samples of 16 Scottish and 26 Swiss sites to test the relationship between reconstructed trait composition from pollen data and vegetation trait composition.

@@

### Vegetation data

@@

### Trait data

Trait data was collected in the middle ring. For determining plant height 10 flowering of individuals were selected at every site if present, non-flowering individuals were selected if flowering individuals were not available. Five leaf samples per species for specific leaf area (SLA), leaf area (LA), leaf dry matter content (LDMC), leaf carbon and leaf nitrogen were collected in sealable plastic bags. Fresh weight and surface area were measured within 72 hours of collection using a carat scale and flatbed scanner. Samples were dried at 70 C for 72 hours before measuring dry weight. For measuring leaf carbon and nitrogen samples using ICP-MS were crushed and X.

@@

### Pollen data

@@

-   Pollen were divided in to plant functional groups according to [@Reitalu_2019] trees shrubs (including Ericaceae family??), and herb (including grasses)

For reconstructing plant trait composition from pollen data, assumptions need to be made about the species belonging to the pollen taxa in the data set. We used the *rgbif* package to download the current distribution of species belonging to each pollen taxon in Scotland and Switzerland from the Global Biodiversity Information Facility database (REF). Species distribution data was queried using the pollen taxonomic name

@@

Correlations were calculated for all zones and three traits (currently): SLA, LA and plant height. Furthermore, four different "treatments" were assigned to test what factors could influence representation of trait composition by pollen data: the use of correction factors, the separation of vegetation and pollen data in to plant functional type and pollination mode. Lastly, the vegetation was aggregated to genus level and family level to test the effect of taxonomic resolution on the reconstructions.

@@

## ![](Figures/veg-Family.png)

## Discussion

-   representativeness of trait databases for trait measurements in the field?

    -   Sandel (2015) - trait databases are biased towards big leaved and big seeded plants
    -   Prinzing (2008) - Less lineages -- more trait variation

-   Comparison between Switzerland and Scotland methods of sampling

-   Some pollen types consider many more species than others for example in this study XX . This is also indentified as a problem in plant richness studies (Reitalu 2019) and trait rep (Carvalho, Blaus). although bayesian xxx

-   size of the relevant source area (Bunting 2003)

## Appendix

@@

### Site characteristics

insert pollen plot

```{r}
# source("03_pollen_diagrams.R")
```

```{r}
source("01_PFT_table_vegetation.R")
sum
```
